**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**

### 2023-02-08 data_cleaning/Data_cleaning_final.qmd
Fixed appendix, cleaning folder, latex preamble

@@

# Preliminary Analyses

This document focuses on exploratory analyses, covariate selection, and outlier identification and removal. The goal of the exploratory analyses was to clean the dataset for formal analyses.

As way of comparison, I used the 2016 census data for Ontario.

@@

The next step consisted in identify the missing rates of the covariates, and determinewhich of those could be included in the model. The following code chunk creates a table with the number of missing observations and percentages.

@@

At this stage in the analysis, we identified that outliers were present in the data. Specifically, individuals that were under 25 years of age and that reported having an income \>110k while living in a household of 1. We therefore, created some diagnostic plots to explore the reported household composition of those \<25 with income\>110k, and we removed the outliers from the dataset.

```{r,outliers}
outliers<-clean_data %>%
  filter(age_group=="16_24" &
           h_size==1 &
           income=="over_110000"

@@

## Race and Ethnicity

Next, we explored how the race and ethnicity information from the dataset compared to the data from the Census. The following chunk creates a summary table for the Race variable.

```{r,ethnic information}
#| echo: false
#| warning: false
#| tbl-cap: Ethnic information from the clean dataset
  t6<-clean_data%>%
    select(race)%>%
    group_by(race)%>%
    summarise(observations = n()) %>%
  mutate(percentage = observations / sum(observations))%>%
    gt()%>%
      fmt_percent(
   columns=percentage,
    decimals=1)

t6
```

It is important to mention that the categories for race/ethnicity provided in the survey did not match the categories used in the Census. Therefore, we used a combination of sources to obtain the information presented in @tbl-races, where the categories used could be matched to the Census Information. The data sources for the table were:

- Fact sheet from the Provice of Ontario for Visible Minorities [link](https://www.ontario.ca/document/2016-census-highlights/fact-sheet-9-ethnic-origin-and-visible-minorities): Used to obtain percentages for Arab, Black, East Asian/Pacific Islander (adding Chinese, Korean, and Japanese percentages), Latin American, Mixed (using the percentage for "multiple visible minority"), Other (obtained by adding the Southeast Asian, Filipino, West Asian, and Minority not identified elsewhere percentages), South Asian. _Accessed on January 05, 2022_

- Census Profile for Ontario [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Aboriginal%20peoples&TABID=1&type=1): Used to obtain percentage of Aboriginal population. _Accessed on January 05, 2022_

- Wikipedia entry for Ontario demographics [link](https://en.wikipedia.org/wiki/Demographics_of_Ontario): To corroborate that the percentage of population reported as "European" in this website matched the percentage obtained for "White Caucasian" that was independently obtained by obtaining the difference in population proportion after subtracting the sum of Visible Minorities and Aboriginal Population percentages.

| Ethnicity/Race | percentage |
|----------------|:-----------|
| Arab           | 1.6%       |
| Black          | 4.7%       |
| East Asian/Pacific Islander     | 6.6%       |
| Indigenous     | 2.8%       |
| Latin American | 1.5%       |
| Mixed          | 1.0%       |
| Other          | 5.2%       |
| South Asian    | 8.7%       |
| White Caucasian| 67.8%      |
| **Total**      | 99.9%      |
: Reference Data from Race/Ethnicity {#tbl-races}

Visible Minorities: 29.3%
Not a Visible Minority: 70.7% (Aboriginal 2.8%, White Caucasian 67.8%)

## Age groups

According to the Census, the distribution of the different the age groups for the province of Ontario are as follows:

| Group age   | Percentage of population |
|-------------|:-------------------------|
| 15-24       | 12.7%                    |
| 25-34       | 12.9%                    |
| 35-44       | 12.8%                    |
| 45-54       | 14.9%                    |
| 55-64       | 13.7%                    |
| 65 and over | 16.7                     |

: Age distributions from the 2016 Census for Ontario

To compare the survey data to the Census, the next chunk creates a barplot of the age groups in the survey. Here, it can be noticed that the age-group distribution from the survey data is different from the Census data.

```{r,trends-by-age}
#| echo: false
#| warning: false
#| fig-height: 4
#| fig-width: 8
#| fig-cap: Age group distributions from the dataset

p2<-clean_data%>%
  select(age_group)%>%
   mutate(age_group_ord = fct_relevel(age_group,
                                      "16_24",
                                      "25_34",
                                      "35_44",
                                      "45_54",
                                      "55_64",
                                  "65_and_over"
                                          ))

  ggplot(p2%>%count(age_group_ord) %>%    # G count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(age_group_ord, n, fill=age_group_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm1

## Income

Survey respondents answered the question "What is your household annual income?". To compare the distribution of responses to this question from the survey and the Census data, we used the "Household total income groups in 2015 for private households" from the Census data (available in the Census Data for Ontario website [link](https://www12.statcan.gc.ca/census-recensement/2016/dp-pd/prof/details/page.cfm?Lang=E&Geo1=PR&Code1=35&Geo2=PR&Code2=01&SearchText=Ontario&SearchType=Begins&SearchPR=01&B1=Income&TABID=1&type=1)).

| Household income range (CAD) | Percentage |
|------------------------------|:-----------|
| \< 15,000                    | 5.7%       |
| 15,000 - 24,999              | 7.5%       |
| 25,000 - 39,999              | 11.6%      |
| 40,000- 59,999               | 15.4%      |
| 60,000 - 89,999              | 19.5%      |
| \>90,000                     | 40.3%      |

: Income percentages from the 2016 Census for Ontario

One difference that is noticeable is that the brackets for income in the census data are different than the brackets used in the survey. The census does CAD 4,999 brackets (e.g., CAD 5,000- CAD 9,9999) up to CAD 49,999, followed by CAD 9,999 brackets up to CAD 99,999. After that, the brackets increase to CAD 24,999, and therefore, it is not possible to obtain percentages for the 90,000-109,999 and \>110,000 brackets from the survey.

Therefore, in the following code chunk, we created an additional category for income in the dataset that matched the information from the Census, and barplots to visualize the proportion of each income bracket in the dataset.

```{r,income}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4

clean_data<-clean_data%>%
    mutate(income_ord = case_when(
                                  income %in% "90000_109999" ~ "over_90000",
                                  income %in%"over_110000" ~ "over_90000",
                                  TRUE ~ as.character(income)
                                ))%>%
  mutate(income_ord=as.factor(income_ord))

p3<-clean_data%>%
  mutate(income_ord = fct_relevel(income_ord,
                                  "under_15000",
                                  "15000_24999",
                                  "25000_39999",
                                  "40000_59999",
                                  "60000_89999",
                                  "over_90000"
                                ))

ggplot(p3%>% count(income_ord)%>%  # count number in each group
         mutate(pct=n/sum(n),               # Calculate percent within each group
                ypos = 200),  # Calculate label positions
       aes(income_ord, n, fill=income_ord)) +
  geom_bar(stat="identity") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct*100),"%"), y=ypos))+
  theme_classic()+
  thm2

We identified differences again between the survey data and the Census data. For example, the \<15,000 bracket accounts for about 25% of the responses, a much higher rate than the proportion in the Census. The other brackets were different as well from the Census data.

@@

```{r,join-datasets}

@@

```{r, missing-geog-area}

@@

```{r,merge-datasets-2}
#load missing municipalities dataset

@@

```{r,get-health-data}

@@

```{r,combine}

@@

```{r,merge-health-regions}

>>>>

# Purpose

This document focuses on the cleaning of the data from the Survey of COVID-19 related Behaviours and Attitudes which was used as the data source for the analysis in the main paper. The cleaning process encompassed exploratory analyses, covariate selection, removal of outliers, and adding geographical information to the data for analysis (Health Regions in Ontario). At the end of the data cleaning process, a dataset ready for for formal analyses was produced. This document can be rendered to re-create all the steps used in the cleaning process and to generate the same dataset that was used for analysis.

Note, however, that code chunks where intermediate csv files are generated (such as those for missing municipalities, or missing Health Regions) are commented in the current document as they were executed once to create the files. Also, to reduce the lenght of the document, the code in the code chunks does not appear in the rendered file.

@@

@@

The next step consisted in identify the missing rates of the covariates, and determine which of those could be included in the model. The following code chunk creates a table with the number of missing observations and percentages.

@@

At this stage in the analysis, we examined the data for outliers. We identified certain entries in the survey that corresponded to individuals that were under 25 years of age and that reported having an income \>110k while living in a household 1. The next code chunk creates a plot of the number of observations that have these characteristics, which were around 20. Next, these outliers were removed from the dataset.

```{r,outliers, echo=FALSE, warning=FALSE,message=FALSE}
outliers<-clean_data %>%
  filter(age_group=="16_24" &
           h_size==1 &
           income=="over_110000"

@@

@@

```{r,join-datasets, echo=FALSE}

@@

```{r, missing-geog-area, echo= FALSE}

@@

```{r,merge-datasets-2, echo=FALSE}
#load missing municipalities dataset

@@

```{r,get-health-data, echo=FALSE}

@@

```{r,combine, echo=FALSE}

@@

```{r,merge-health-regions, echo=FALSE}

### 2023-02-08 manuscript/02-methods.qmd
Fixed appendix, cleaning folder, latex preamble

The original dataset contained 39,029 entries (where each entry corresponded to a set of answers provided by a unique respondent). Following a preliminary analysis, the dataset was cleaned in order to only contain the socio-economic information provided in @tbl-covariates and vaccination status. The cleaning process also included removing outliers that were identified during the preliminary analyses. Specifically, we removed those respondents that indicated to be below 25 years of age, living in a household of size 1, and that reported an income above CAD 110,000. After cleaning, the dataset contained 3,709 unique entries.

@@

For each survey participant certain data was automatically captured, including the nearest municipality (city). This resulted in a total of 578 different municipalities within the clean dataset. Due to our interest in analyzing the differences between Health Regions, we assigned the city of each entry to its correspondent Health Region following a multi-step process. Briefly, we used Local Health Integrated Networks (LHINs) to assign a Health Region to each entry in the survey. LHINs were the geographical divisions for health used by  Ontario before the adoption of the Health Regions; because of the lack of a publicly available list of all municipalities within each Health Region, we used a dataset of long-term care homes and LHINs to match each city to LHIN, followed by matching each LHIN to a Health Region following the information provided on the Ontario Health Website, where the list of LHINs and corresponding Health Regions is available. In the case of municipalities that did not appear in the long-term dataset, we manually searched each city in the LHINs websites in order to provide geographical information. Details of the data cleaning process are in the github repository: https://github.com/aimundo/Fields_COVID-19/.

>>>>

The original dataset contained 39,029 entries (where each entry corresponded to a set of answers provided by a unique respondent). Following a preliminary analysis, the dataset was cleaned in order to only contain the socio-economic information provided in @tbl-covariates and vaccination status. The cleaning process also included removing outliers that were identified during the preliminary analyses. Specifically, we removed those respondents that indicated to be below 25 years of age, living in a household of size 1, and that reported an income above CAD 110,000. The next step consisted in processing the geographical information in the survey (city where the survey was responded) in order to match each city to its correspondent Health Region. Details of this process are given below.

@@

For each survey participant certain data was automatically captured, including the nearest municipality (city). This resulted in a total of 578 different municipalities within the clean dataset. Due to our interest in analyzing the differences between Health Regions, we assigned the city of each entry to its correspondent Health Region following a multi-step process. Briefly, we used Local Health Integrated Networks (LHINs) to assign a Health Region to each entry in the survey. LHINs were the geographical divisions for health used by  Ontario before the adoption of the Health Regions; because of the lack of a publicly available list of all municipalities within each Health Region, we used a dataset of long-term care homes and LHINs to match each city to LHIN, followed by matching each LHIN to a Health Region following the information provided on the Ontario Health Website, where the list of LHINs and corresponding Health Regions is available. In the case of municipalities that did not appear in the long-term dataset, we manually searched each city in the LHINs websites in order to provide geographical information. The raw dataset, clean dataset, details of the data cleaning process, and the addition of Health Region and LHIN information are described in detail in the github repository: https://github.com/aimundo/Fields_COVID-19/.

### 2023-02-08 manuscript/03-results.qmd
Fixed appendix, cleaning folder, latex preamble

```{r,model-table}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-model
#| tbl-cap: "Multiple Regression Analysis-Predictors of Vaccination Status"
source(here::here("code","raking_logistic_regression.R"))

m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```

>>>>

```{r,model-table}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-model
#| tbl-cap: "Multiple Regression Analysis-Predictors of Vaccination Status"

#logistic regression
source(here::here("code","regression.R"))

# create table with the model output
m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```
